Formalizing Stack Safety as a Security Property

### 2022-12-04 submission5/noexceptions.tex
More detail at the end, and fixed formatting

Allocating public objects affects the call stack very similarly to private ones,
but instead of labeling them \(\object\) they become \(\public\), so they are
not sealed at a call. This makes these
regions of memory able to transmit information between caller and callee without
violating the stack safety properties. Note that there is no way to seal a location
once it has become \(\public\); only when {\tt x} is deallocated will its region
be available to allocate privately.

\judgmentbr[AllocPriv]
           {\(b = \mach[\SP] + \mathit{off}\)}
           {\(V' = V \llbracket \addr \mapsto \sealed |
             b \leq a < b+\mathit{sz} \land V ~ \addr = \unsealed \rrbracket\)}
           {\(Op ~ \mach ~ (\mathbf{alloc} ~ \mathbf{f} ~ \mathit{off, sz}) ~ (V,\sigma) = (V',\sigma)\)}

\judgmentbr[AllocPub]
           {\(b = \mach[\SP] + \mathit{off}\)}
           {\(V' = V \llbracket \addr \mapsto \public |
             b \leq a < b+\mathit{sz} \land V ~ \addr = \unsealed \rrbracket\)}
           {\(Op ~ \mach ~ (\mathbf{alloc} ~ \mathbf{t} ~ \mathit{off, sz}) ~ (V,\sigma) = (V',\sigma)\)}

\judgmentbr[Dealloc]
           {\(b = \mach[\SP] + \mathit{off}\)}
           {\(V' = V \llbracket \addr \mapsto \unsealed |
             b \leq a < b+\mathit{sz} \land V ~ \addr \in \{\object, \public\} \rrbracket\)}
           {\(Op ~ \mach ~ (\mathbf{dealloc} ~ \mathit{off, sz}) ~ (V,\sigma) = (V',\sigma)\)}

@@

So we must refine our call operation to make use of the information that we have about
which memory contain arguments, \(\overline{sa}\). \(\overline{sa}\) is a set of
triples of a register, an offset from the value of that register, and a size.
We first define the helpful set \(\mathit{passed} ~ \overline{sa} ~ \mach\),
then extend the call operation to keep all objects in \(\mathit{passed}\) marked
as \(\object\) and seal everything else.
%
\[\mathit{passed} ~ \overline{sa} ~ \mach \triangleq
\bigcup_{(\reg,b,o) \in \overline{sa}} \{\mach[\reg]+i | b \leq i < b+o\}\]
%
\judgmentbrbr[]
             {\(\psi = \mathbf{call} ~ \addr_{target} ~ \overline{\reg_{args}} ~ \overline{sa}\)}
             {\(V' = V \llbracket \reg \mapsto \unsealed | \reg \in \mathit{CLR} \rrbracket
               \llbracket \reg \mapsto \public | \reg \in \overline{\reg_{args}} \rrbracket\)}
             {\(V'' = V'\llbracket \addr \mapsto \sealed | V' ~ \addr = \object \land \addr \not \in (\mathit{passed} ~ \overline{sa} ~ \mach) \rrbracket\)}
             {\(Op ~ \mach ~ \psi ~ (V,\sigma) =
               (V'',V::\sigma)\)}

@@

Since the caller will not be returned to, it does not need integrity, though
it should still enjoy confidentiality. We set its frame to \(\unsealed\) rather
than \(\sealed\) to express this. Of course, some objects may be passed in place,
easily modeled by including them in \(\overline{sa}\).

@@

We can model this as a pair of additional security-relevant operations: one which
declares a register to contain a valid pointer, and one which transmits the provenance
of a pointer from one element to another.

\begin{align*}
  \psi \in \Psi ::= & \dots \\
  & \mathbf{promote} ~ \reg ~ sa & sa \in \REGS \times \mathbb{Z} \times \mathbb{N} \\
  & \mathbf{propagate} ~ \component_{src} ~ \component_{dst} \\
\end{align*}

@@

\judgmenttwobrlong[]
                  {\(\psi = \mathbf{promote} ~ \reg_{dst} ~ (\reg_{base},\mathit{off},sz)\)}
                  {\(b = \mach[\reg_{base}]+\mathit{off}\)}
                  {\(\rho' = \rho[\reg \mapsto (b,b+sz)]\)}
                  {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) = (V,\sigma,\rho')\)}

\judgmenttwo[]
            {\(\psi = \mathbf{propagate} ~ \component_{src} ~ \component_{dst}\)}
            {\(\rho' = \rho[\component_{dst} \mapsto \rho[\component_{src}]]\)}
            {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) = (V,\sigma,\rho')\)}

We now have a notion of provenance, and must integrate it into the definition of
stack safety. We essentially generalize the above notion of passing: we will consider
a caller to have intentionally passed an object if that object is reachable by
a capability that has been passed to the callee. This includes capabilities passed
indirectly, by being stored in an object that is in turn passed. Formally, we call
this set \(\mathit{capped}\), and define it recursively:

\[\begin{split}
& \mathit{capped} ~ \components ~ \rho \triangleq \bigcup_{\component \in \components} \{\component\} \cup \mathit{capped} ~ \components' ~ \rho \textnormal{ where} \\
& \components' = \{\component' |\rho[\component] = (\mathit{base},\mathit{bound})
\land \mathit{base} \leq \component' < \mathit{bound}\} \\
\end{split}\]

@@

>>>>

Allocating public objects affects the call stack very similarly to private ones,
but instead of labeling them \(\object\) they become \(\public\), so they are
not sealed at a call. This makes these
regions of memory able to transmit information between caller and callee without
violating the stack safety properties. Note that there is no way to seal a location
once it has become \(\public\); only when {\tt x} is deallocated will its region
be available to allocate privately. We define a helper function, \(\mathit{range}\),
to map an offset from a register and a size into a set of addresses.
%
\[\mathit{range} ~ \reg ~ \mathit{off} ~ \mathit{sz} ~ \mach \triangleq
\{\mach[\reg]+i | \mathit{off} \leq i < \mathit{off+sz}\}\]
%
\judgmentbr[AllocPriv]
           {\(\components = \mathit{range} ~ \SP ~ \mathit{off} ~ \mathit{sz} ~ \mach \cap
             \{\addr \mid V ~ \addr = \unsealed\}\)}
           {\(V' = V \llbracket \addr \mapsto \sealed \mid \addr \in \components \rrbracket\)}
           {\(Op ~ \mach ~ (\mathbf{alloc} ~ \mathbf{f} ~ \mathit{off, sz}) ~ (V,\sigma) = (V',\sigma)\)}

\judgmentbr[AllocPub]
           {\(\components = \mathit{range} ~ \SP ~ \mathit{off} ~ \mathit{sz} ~ \mach \cap
             \{\addr \mid V ~ \addr = \unsealed\}\)}
           {\(V' = V \llbracket \addr \mapsto \public \mid \addr \in \components \rrbracket\)}
           {\(Op ~ \mach ~ (\mathbf{alloc} ~ \mathbf{t} ~ \mathit{off, sz}) ~ (V,\sigma) = (V',\sigma)\)}

\judgmentbr[Dealloc]
           {\(\components = \mathit{range} ~ \SP ~ \mathit{off} ~ \mathit{sz} ~ \mach \cap
             \{\addr \mid V ~ \addr = \unsealed\}\)}
           {\(V' = V \llbracket \addr \mapsto \unsealed \mid \addr \in \components \rrbracket\)}
           {\(Op ~ \mach ~ (\mathbf{dealloc} ~ \mathit{off, sz}) ~ (V,\sigma) = (V',\sigma)\)}

@@

So we must refine our call operation to make use of the information that we have about
which memory contain arguments, \(\overline{sa}\). \(\overline{sa}\) is a set of
triples of a register, an offset from the value of that register, and a size.
We first define the helpful set \(\mathit{passed} ~ \overline{sa} ~ \mach\),
then extend the call operation to keep all objects in \(\mathit{passed}\) marked
as \(\object\) and seal everything else.
%
\[\mathit{passed} ~ \overline{sa} ~ \mach = \bigcup_{(\reg,\mathit{off},\mathit{sz}) \in \overline{sa}}
\mathit{range} ~ \reg ~ \mathit{off} ~ \mathit{sz} ~ \mach\]
%
\judgmentbrbr[]
             {\(\psi = \mathbf{call} ~ \addr_{target} ~ \overline{\reg_{args}} ~ \overline{sa}\)}
             {\(V' = V \llbracket \reg \mapsto \unsealed | \reg \in \mathit{CLR} \rrbracket
               \llbracket \reg \mapsto \public | \reg \in \overline{\reg_{args}} \rrbracket\)}
             {\(V'' = V'\llbracket \addr \mapsto \sealed | V' ~ \addr = \object \land \addr \not \in (\mathit{passed} ~ \overline{sa} ~ \mach) \rrbracket\)}
             {\(Op ~ \mach ~ \psi ~ (V,\sigma) =
               (V'',V::\sigma)\)}

@@

Since the caller will not be returned to, it does not need integrity, though
it should still enjoy confidentiality. We set its frame to \(\unsealed\) rather
than \(\sealed\) to express this. Of course, some objects may be passed as arguments
in place, easily modeled by including them in \(\overline{sa}\). More sophisticated
behaviors may require sequences of {\bf dealloc}, {\bf tailcall}, then {\bf alloc}.

@@

We can model this as a trio of additional security-relevant operations: one which
declares a register to contain a valid pointer, one which transmits the provenance
of a pointer from one element to another, and one which clears the provenance
(for instance, when a pointer is modified in place in a way that makes it invalid.)

\begin{align*}
  \psi \in \Psi ::= & \dots \\
  & \mathbf{promote} ~ \reg ~ sa & sa \in \REGS \times \mathbb{Z} \times \mathbb{N} \\
  & \mathbf{propagate} ~ \component_{src} ~ \component_{dst} \\
  & \mathbf{clear} ~ \component \\
\end{align*}

@@

\judgmentbr[Promote]
           {\(\psi = \mathbf{promote} ~ \reg_{dst} ~ (\reg_{base},\mathit{off},\mathit{sz})\)}
           {\(\rho' = \rho[\reg_{dst} \mapsto \mathit{range} ~ \reg_{base} ~ \mathit{off} ~ \mathit{sz}]\)}
           {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) = (V,\sigma,\rho')\)}

\judgmentbr[Propagate]
           {\(\psi = \mathbf{propagate} ~ \component_{src} ~ \component_{dst}\)}
           {\(\rho' = \rho[\component_{dst} \mapsto \rho[\component_{src}]]\)}
           {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) = (V,\sigma,\rho')\)}

\judgment[Clear]
         {\(\psi = \mathbf{clear} ~ \component\)}
         {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) = (V,\sigma,\rho[\component \mapsto \emptyset])\)}

We now have a notion of provenance, and must integrate it into the definition of
stack safety. We essentially generalize the above notion of passing: we will consider
a caller to have intentionally passed an object if that object is reachable by
a capability that has been passed to the callee. This includes capabilities passed
indirectly, by being stored in an object that is in turn passed. Formally, we call
this set \(\mathit{capped}\), and define it recursively:
%
\[\begin{split}
& \mathit{capped} ~ \components ~ \rho \triangleq \bigcup_{\component \in \components} \{\component\} \cup \mathit{capped} ~ \components' ~ \rho \textnormal{ where} \\
& \components' = \{\component' |\rho[\component] = (\mathit{base},\mathit{bound})
\land \mathit{base} \leq \component' < \mathit{bound}\} \\
\end{split}\]

@@

There are some advanced features that would be nice to work in. Any of these that we support should
all be folded into the ``advanced machine,'' so that the full formalism is only presented once.
They can still be explained progressively. First, concurrency: the coroutine properties from earlier
drafts should translate just fine into this formalism, but the challenge is that concurrency introduces
non-determinism. I think it's feasible to formalize. I do wonder what level of testing is appropriate;
two threads switching at intervals and mostly minding their own business seems doable, but the interesting
parts happen when they interact through memory.

It would also be nice to have a story about exceptions. I've done some thinking about how to model them,
and we'll want to choose one particular style (e.g., C++-scoping with or without stack unwinding).
